On Optimality of Finite Element Integration

### 2015-10-18 TOMS_Paper/assembly.tex
Trying to rewrite algorithm...

\usepackage{listings}
\lstset{language=C, breaklines=true, mathescape,
			 morekeywords={and, each}}

@@

\section{Problem Characterization}
\label{sec:optimal-impl}

@@

\subsection{Transformation Space - Sharing Elimination}
\label{sec:sharing-elimination}
We start with introducing the fundamental notion of sharing.

@@

\subsection{Transformation Space - Pre-evaluation}
\label{sec:pre-evaluation}
For finite element integration loop nests, the presence of sharing is a \textit{sufficient but not necessary} condition for being in \textit{non-optimal} form. Consider again the bilinear form implementation in Figure~\ref{code:loopnest}. We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction imposed by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $M$? The transformation we look for is exemplified in Figures~\ref{code:loopnest_red} and~\ref{code:loopnest_nored}; Figure~\ref{code:loopnest_red}, which represents the input, can be seen as a simple instance of the abstract loop nest in Figure~\ref{code:loopnest}.

@@

\subsection{Completeness of the Transformation Space}
It is helpful to consider sharing elimination and pre-evaluation as high level operators defined on top of basic transformations such as code motion, expansion, and factorization. These two operators allow us to analyze all possible ways in which an expression can be factorized; that is, by grouping:
\begin{itemize}
\item Test or trial functions (sharing elimination in the outer product loop nest)
\item Integration-dependent terms (sharing elimination in the reduction loop)
\item Together test functions, trial functions, and integration-dependent terms (pre-evaluation)
\end{itemize}
Factorization addresses \textit{spatial redundancy}. The presence of spatial redundancy means that some operations are needlessly executed at two points in an expression.

@@

Sharing elimination and pre-evaluation allow us to construct an algorithm to explore all ways in which spatial and temporal redundancy can be eliminated from loop nests and expressions as in Definition~\ref{def:fem-loopnest}.

\section{Transformation Algorithm}
\label{sec:optimal-synthesis}

In this section, we build a transformation algorithm capable of reducing finite element integration loop nests to optimal form. A fundamental aspect is the definition of a suitable cost function to assess the profitability of pre-evaluation (for which a deeper explanation will be provided). The optimality of the transformation algorithm is discussed.

@@

\subsection{Reaching Optimality}
We here discuss how we can systematically reach the goal set by Definition~\ref{def:optimality} under Constraints~\ref{const:Le} and~\ref{const:TH} for the class of loop nests in Definition~\ref{def:fem-loopnest}.

We reinforce the idea that eliminating sharing from the outer product loop nest does not suffice. As presaged in Section~\ref{sec:optimal-impl}, we consider whether, and under what conditions, the reduction imposed by $L_i$ could be pre-evaluated, thus reducing the operation count.

Two issues should be addressed: understanding for which monomials pre-evaluation is profitable; coordinating pre-evaluation with sharing elimination, since the latter inhibits the former. In this section, we tackle the second issue, while we deal with the first one in Section~\ref{sec:op_count}. For this, we start with explaining the transformation algorithm, for which pseudo-code is provided in Figure~\ref{code:intuition}.

\begin{figure}
\begin{lstlisting}[numbers=left, basicstyle=\small\ttfamily, frame=single]
dissect the input expression into monomials
for each monomial $m$:
  if $\theta(m)$ > 0 and memory constraints satisfied:
    extract into a separate loop nest
    apply pre-evaluation
for each expression:
  remove sharing    
\end{lstlisting}
\caption{High level view of the transformation algorithm}
\label{code:intuition}
\end{figure}

Our choice is to evaluate the impact of pre-evaluation, as number of operations saved or introduced, ``locally''; that is, for each monomial, ``in isolation''. We then introduce the cost function $\theta : M \rightarrow \mathbb{Z}$, where $M$ is the set of monomials in the form. $\theta(m)$ represents the change in operation count if pre-evaluation were applied to monomial $m$: a negative value indicates an increase, whereas a positive value indicates a gain. If pre-evaluation is found profitable (i.e., $\theta(m) > 0$), the sub-expression implementing $m$ is extracted; the surrounding loop nest is suitably ``cloned''; a sequence of transformation steps -- involving expansion, factorization, and code motion -- takes place (details in Section~\ref{sec:codegen}); and, finally, the hoisted code is pre-evaluated using symbolic execution. The result is a set of $n$-dimensional tables (these can be seen as ``slices'' of the reference tensor at the math level), $n$ being the arity of the multilinear form.

>>>>

\usepackage{listings}
\lstset{language=C, breaklines=true, mathescape,
			 morekeywords={each}}

@@

\section{Transformation Space}
\label{sec:optimal-impl}

@@

\subsection{Sharing Elimination}
\label{sec:sharing-elimination}
We start with introducing the fundamental notion of sharing.

@@

\subsection{Pre-evaluation}
\label{sec:pre-evaluation}
For finite element integration loop nests, the presence of sharing is a \textit{sufficient but not necessary} condition for being in \textit{non-optimal} form. Consider again the bilinear form implementation in Figure~\ref{code:loopnest}. We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction imposed by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $M$? The transformation we look for is exemplified in Figures~\ref{code:loopnest_red} and~\ref{code:loopnest_nored}; Figure~\ref{code:loopnest_red}, which represents the input, can be seen as a simple instance of the abstract loop nest in Figure~\ref{code:loopnest}.

@@

\subsection{Completeness}
It is helpful to consider sharing elimination and pre-evaluation as high level operators defined on top of basic transformations such as code motion, expansion, and factorization. These two operators allow us to analyze all possible ways in which an expression can be factorized; that is, by grouping:
\begin{itemize}
\item Test or trial functions (sharing elimination in the outer product loop nest)
\item Integration-dependent terms (sharing elimination in the reduction loop)
\item Together test functions, trial functions, and integration-dependent terms (pre-evaluation)
\end{itemize}
Factorization addresses \textit{spatial redundancy}. The presence of spatial redundancy means that some operations are needlessly executed at two points in an expression.

@@

Sharing elimination and pre-evaluation allow us to construct an algorithm to minimize spatial and temporal redundancy in loop nests and expressions as in Definition~\ref{def:fem-loopnest}.

\section{Framework for Optimal Synthesis}
\label{sec:optimal-synthesis}

In this section, we build a transformation algorithm capable of reducing finite element integration loop nests to optimal form. A fundamental aspect is the definition of a suitable cost function to assess the profitability of applying pre-evaluation to a monomial.

@@

\subsection{Transformation Algorithm}
We here discuss how we can systematically reach the goal set by Definition~\ref{def:optimality} under Constraints~\ref{const:Le} and~\ref{const:TH} for the class of loop nests in Definition~\ref{def:fem-loopnest}.

Two main issues should be addressed: 
\begin{itemize}
\item Understanding for which monomials pre-evaluation is profitable. In order to expose pre-evaluation opportunities, Algorithm~\ref{algo:pre-evaluation} transforms the input expression by expanding products. This increases the operation count. We therefore say that pre-evaluation is profitable if the gain due to pre-evaluating the reduction loop is larger than such increase.
\item Consider a form comprising two monomials $m1$ and $m2$. Assume that pre-evaluation is profitable for $m1$ but not for $m2$, and that $m1$ and $m2$ share at least one term (e.g. some basis functions). If pre-evaluation were applied to $m1$, sharing between $m1$ and $m2$ would be lost. We then need a mechanism to understand what transformation -- pre-evaluation or sharing elimination -- provides the highest operation count reduction.
\end{itemize}

For the moment, let us assume that we own a cost function $\theta : M \rightarrow \mathbb{Z}$ that, given a monomial $m \in M$, returns the number of operations saved or introduced if pre-evaluation were applied to $m$. $\theta(m) < 0$ indicates that pre-evaluation is not profitable, because the operation count would actually increase by $|\theta(m)|$. Analogous considerations stand for $\theta(m) > 0$. We will show how to construct a proper $\theta$ in Section~\ref{sec:op_count}. Having defined $\theta$, we can now elaborate on the transformation algorithm, for which pseudo-code is provided in Figure~\ref{code:intuition}.

\begin{figure}
\begin{lstlisting}[numbers=left, basicstyle=\small\ttfamily, frame=single]
M = $\lbrace$m | m is a monomial in the input expression$\rbrace$
S = $\lbrace$m | $\theta(m)$ < 0$\rbrace$
P = $\lbrace$m | $\theta(m)$ $\geq$ 0 and sharing(m, S) = $\emptyset \rbrace$
maybe_P = M $\setminus$ (S $\cup$ P)
B = $\lbrace$b | b is a bipartite graph from complete graph G = (maybe_P, E)$\rbrace$
for each $b \in B$:
  for each $(b_S, b_P) \in b$:

for each expression:
  remove sharing    
\end{lstlisting}
\caption{High level view of the transformation algorithm}
\label{code:intuition}
\end{figure}

If pre-evaluation is found profitable (i.e., $\theta(m) > 0$), the sub-expression implementing $m$ is extracted; the surrounding loop nest is suitably ``cloned''; a sequence of transformation steps -- involving expansion, factorization, and code motion -- takes place (details in Section~\ref{sec:codegen}); and, finally, the hoisted code is pre-evaluated using symbolic execution. The result is a set of $n$-dimensional tables (these can be seen as ``slices'' of the reference tensor at the math level), $n$ being the arity of the multilinear form.
